---
title: "Building the conceptual framework"
subtitle: "Identify the goals"
---

```{r data}
#| echo: false
#| warning: false
#| message: false

source("../_data/functions/start_script.R")

```


One of the first steps to building the conceptual OHI framework is determining which goals to include.

The global OHI assessment includes 10 goals. But some of these goals may not be relevant to the Gulf of California. Conversely, there may be goals that are important to the region that were not included in the global assessment.

And, even when goals are the same, the models and data underlying the goal will very likely be different.

There is no reason to follow the global model, but understanding the theory behind the global goals and how they were measured is a helpful way to begin thinking about what to include in the Gulf of California assessment. 

<br>

::: {.callout-note collapse=true}

## Learn more about the global goals (click to expand)

Clicking on the goal will link to the goal pages which provide more information about the **global** goals.
```{r}
#| echo: false
#| warning: false
#| message: false

goals_df  <- read_csv("../_data/output/goals.csv") 
goals_df$subgoal <- sapply(goals_df$subgoal, function(x) 
  if (!is.na(x)) paste0(strrep('&nbsp;', 4), x) else x)

  datatable(goals_df, escape=FALSE, rownames=FALSE)
```

:::

Here is a [resource](https://ohi-science.org/toolbox-training/data-planner.html) to help you develop the goals for the Gulf of California.