---
title: "Ocean Health Index Gulf of California"
subtitle: "Current GoC OHI Regions"
toc: false
editor_options: 
  chunk_output_type: console
---


```{r}
#| echo: false
#| warning: false
#| message: false
#| output: false
#| results: hide

library(magrittr)
library(mapview)
library(sf)
library(tidyverse)
library(here)

# read in data
ohi_nolines_git <- st_read("https://raw.githubusercontent.com/OHI-GOC/ohi_goc_prep/refs/heads/gh-pages/spatial/marine_ohi/ohirgns_nolines.geojson", quiet=TRUE)

## for plotting:
rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)
```

```{r, fig.width = 10, fig.height = 10}
#| echo: false
#| warning: false
#| message: false


mapview(ohi_nolines_git, zcol = "rgn_id",layer.name = "GoC OHI regions", col.regions = rgn_colors) 
```

<div class="dark-table">

| Region | English name | Nombre en español |
|--------|--------------|------------------|
| Region 1 | Western Upper Gulf | Alto Golfo Occidental |
| Region 2 | Western Midriff Islands | Islas Grandes Occidentales  |
| Region 3 | Western Central Gulf | Golfo Central Occidental |
| Region 4 | Western Southern Gulf | Golfo Sur Occidental |
| Region 5 | Eastern Upper Gulf | Alto Golfo Oriental |
| Region 6 | Eastern Midriff Islands |  Islas Grandes Orientales |
| Region 7 | Southern Sonora Coast | Costa Sur de Sonora |
| Region 8 | Sinaloa Coast | Costa de Sinaloa |
| Region 9 | Southeastern Gulf |  Golfo Sur Oriental |

</div>

<style>
.dark-table table {
  border-collapse: collapse;
}

.dark-table th,
.dark-table td {
  border: 0.5px solid #444;
}
</style>



