---
title: "Overall Update"
format: html
toc: true
---

We are excited to share where the Gulf of California Ocean Health Index currently stands. Expert Working Groups 1, 2, and 3 (including 3.1 - 3.3) are now complete, and the plots below show each goal’s current status by OHI region. If you would like, you can hover your mouse over each plot to compare regions and years. A score of 100 means that the current status is equal to the reference point.

This is a snapshot of work in progress. Scores, methods, and write-ups will continue to be updated as we finish pressures (including Safety & Security), finalize the goals, and complete the resilience layers. Specific data sources and reference points for each goal are described in detail in that goal’s executive summary.

| Date | OHI assessment | Paper |
|------|----------------|-------|
| September 30 | Finish pressures (including Safety & Security) and finalize goals | Finalize goals / work on paper |
| October 30 | Finish resilience | Work on paper |
| November 13 | Send OHI draft to the core team and co-authors | Work on paper |
| December 4 | Deadline for all co-authors to provide feedback | Work on paper |
| January 1 | Submit OHI paper | Finalize paper |


```{r}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(plotly)
library(here)

rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

region_levels <- paste0("Region_", 1:9)
rgn_label_colors <- setNames(
  unname(rgn_colors[region_levels]),
  unname(rgn_labels[region_levels])
)

add_rgn_labels <- function(df) {
  df %>%
    mutate(
      rgn_id = as.character(rgn_id),
      rgn_id = stringr::str_replace_all(rgn_id, " ", "_"),
      rgn_id = ifelse(
        stringr::str_detect(rgn_id, "^[0-9]+$"),
        paste0("Region_", rgn_id),
        rgn_id
      ),
      rgn_id = factor(rgn_id, levels = region_levels),
      rgn_label = factor(
        rgn_labels[as.character(rgn_id)],
        levels = unname(rgn_labels[region_levels])
      )
    )
}

prep_cs <- function(df) {
  if ("OHI_region_no" %in% names(df) && !"rgn_id" %in% names(df)) {
    df <- df %>% dplyr::rename(rgn_id = OHI_region_no)
  }

  acronym_score <- names(df)[stringr::str_detect(names(df), "_[Ss]core$")]
  acronym_score <- setdiff(acronym_score, "current_status")
  if (!"current_status" %in% names(df) && length(acronym_score) >= 1) {
    df <- df %>% dplyr::rename(current_status = dplyr::all_of(acronym_score[[1]]))
  }
  if (!"current_status" %in% names(df) && "score" %in% names(df)) {
    df <- df %>% mutate(current_status = score)
  }

  df %>%
    mutate(
      rgn_id = as.character(rgn_id),
      rgn_id = stringr::str_replace(rgn_id, "^region_", "Region_"),
      year = as.numeric(year),
      current_status = as.numeric(current_status)
    ) %>%
    add_rgn_labels()
}

goal_plotly <- function(df, title) {
  df <- df %>%
    arrange(rgn_label, year)

  year_breaks <- sort(unique(df$year[!is.na(df$year)]))
  df_drawn <- df %>% filter(!is.na(current_status))

  plot_ly(
    df_drawn,
    x = ~year,
    y = ~current_status,
    color = ~rgn_label,
    colors = rgn_label_colors,
    type = "scatter",
    mode = "lines+markers",
    connectgaps = TRUE
  ) %>%
    layout(
      title = title,
      xaxis = list(
        title = "Año // Year",
        tickmode = "array",
        tickvals = year_breaks,
        ticktext = as.character(as.integer(year_breaks)),
        range = range(year_breaks)
      ),
      yaxis = list(title = "Puntaje // Percent (out of 100)", range = c(0, 105)),
      legend = list(title = list(text = "Region"))
    ) %>%
    add_lines(
      x = range(year_breaks),
      y = c(100, 100),
      line = list(color = "black", width = 2, dash = "dash"),
      name = "Meta // Goal (100/100)",
      inherit = FALSE
    )
}
```

# ![](figs/tourism-and-recreation.svg){width=8% height=8%} Tourism & Recreation {#tr}

The overall TR score is the average of Quantity, Quality, and Sustainability. A score of 100 is the ideal sustainable state. Open the component scores below to see what is pulling a region up or down. Specific data sources and reference points are described in detail in the executive summary.

```{r}
#| echo: false
#| warning: false
#| message: false

tr_root <- "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025"

tr_prep <- function(df, score_col) {
  df %>%
    mutate(
      year = as.numeric(year),
      current_status = as.numeric(.data[[score_col]])
    ) %>%
    filter(year >= 2019, year <= 2025) %>%
    add_rgn_labels()
}

tr_quantity <- read_csv(
  file.path(tr_root, "output_quantity/tr_quantity_current_status_2019_2026_OGmodel.csv"),
  show_col_types = FALSE
) %>%
  mutate(score = ifelse(score > 100, 100, score)) %>%
  tr_prep("score")

tr_quality_crime <- read_csv(
  file.path(tr_root, "output_quality/tr_qual_crime_score_030626.csv"),
  show_col_types = FALSE
) %>%
  dplyr::select(rgn_id, year = ano, score = current_status) %>%
  mutate(quality_type = "crime")

tr_quality_activities <- read_csv(
  file.path(tr_root, "output_quality/final_activity_avg_score_082826.csv"),
  show_col_types = FALSE
) %>%
  dplyr::select(rgn_id, year, score = final_score) %>%
  mutate(quality_type = "activities")

tr_quality <- bind_rows(tr_quality_activities, tr_quality_crime) %>%
  filter(year %in% 2019:2025) %>%
  pivot_wider(names_from = quality_type, values_from = score) %>%
  mutate(final_score = rowMeans(cbind(activities, crime), na.rm = TRUE)) %>%
  tr_prep("final_score")

tr_sust <- read_csv(
  file.path(tr_root, "output_sustainability/cs_sust_all_indicators_083126.csv"),
  show_col_types = FALSE
) %>%
  tr_prep("final_score")

tr_overall <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/final_cs_scores/TR_Scores_1.0_df_091126.csv"),
  show_col_types = FALSE
) %>%
  prep_cs()

goal_plotly(tr_overall, "Tourism & Recreation current status")
```

- **English:** [TR Executive Summary](https://docs.google.com/document/d/1y8W_z3GJL1SY4x7m_QrXP7D1vvfl53ghAgC0hB6VbD0/edit?tab=t.0)
- **Spanish:** [TR Resumen ejecutivo](https://docs.google.com/document/d/1RhY1pTaZVDXHtm4OTRL5f6lf5fKj8fOLhRi1rCGCN7k/edit?tab=t.0)
- **EWG 3.1 page:** [Tourism & Recreation](ewg31_tr.html)
- **Goalkeeper page:** [Tourism and Recreation](tr.html)

::: {.callout-note collapse="true"}
# Component scores

Specific data sources and reference points for each component are described in detail in the executive summary.

This is the Quantity component, which evaluates whether international and domestic tourism is reaching the targets set by Plan México for 2030.

```{r}
#| echo: false
#| warning: false
#| message: false
goal_plotly(tr_quantity, "TR Quantity")
```

This is the Quality component, which evaluates whether visits are safe and well supported by dining, cultural, and recreation services, relative to the best conditions observed in the Gulf.

```{r}
#| echo: false
#| warning: false
#| message: false
goal_plotly(tr_quality, "TR Quality")
```

This is the Sustainability component, which evaluates whether tourism is supported by certified hotels and beaches, quality jobs, and public investment at or above recent historical levels.

```{r}
#| echo: false
#| warning: false
#| message: false
goal_plotly(tr_sust, "TR Sustainability")
```
:::

::: {.callout-note collapse="true"}
# Next steps and questions for Alejandro

**Next steps (from EWG 3.1)**

- *To be added from EWG feedback.*

**Questions for Alejandro**

- Do these scores seem right to you?
- Do you have any major questions or concerns you would like to discuss?
- Are there any regionally important sustainability certifications missing (currently Distintivo S, Green Globe, Green Key, and Blue Flag)?
:::

# ![](figs/food-provision.svg){width=8% height=8%} Food Provisioning {#fp}

```{r}
#| echo: false
#| warning: false
#| message: false
fp_cs <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/final_cs_scores/FP_Scores_1.0_df_Sep112026.csv"),
  show_col_types = FALSE
) %>%
  prep_cs()
goal_plotly(fp_cs, "Food Provisioning current status")
```

*Sophia will move Erica's Food Provisioning current-status plot into `figs/` before Friday.*

::: {.callout-tip collapse="true"}
# About this goal

The Food Provisioning (FP) goal measures the amount of seafood that is sustainably produced by wild-catch fisheries and mariculture in the Gulf of California. Specific data sources and reference points are described in detail in the executive summary.

- **English:** [FP Executive Summary](https://docs.google.com/document/d/1sYug__6lxXjE_Vg3zW0etdAqZKKU3eJZtiX4bzd7UZA/edit?tab=t.0)
- **Spanish:** [FP Resumen Ejecutivo](https://docs.google.com/document/d/1uvmC99z4ZNi7pv5NdgYGATG8WRFpqUipA0c1H4NqPiw/edit?tab=t.0)
- **EWG 3.1 page:** [Food Provisioning](ewg31_fp.html)
- **Goalkeeper page:** [Food Provisioning](fp.html)

:::

::: {.callout-note collapse="true"}
# Next steps

- *To be added from EWG feedback.*
:::

# ![](figs/livelihoods-and-economies.svg){width=8% height=8%} Livelihoods & Economies {#le}

```{r}
#| echo: false
#| warning: false
#| message: false
le_cs <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/final_cs_scores/LE_Scores_1.0_df_091126.csv"),
  show_col_types = FALSE
) %>%
  prep_cs()
goal_plotly(le_cs, "Livelihoods & Economies current status")
```

::: {.callout-tip collapse="true"}
# About this goal

The Livelihoods & Economies (LE) goal measures jobs and revenue from marine-related industries. Livelihoods (LIV) captures the quantity and quality of marine-related jobs. Economies (ECO) captures revenue from marine-related sectors. Specific data sources and reference points are described in detail in the executive summary.

- **English:** [LE Executive Summary](https://docs.google.com/document/d/1EQd0xKKkfeEm_wsYa8bZMxTqH7OWtk3-BQ-UHAg1LXA/edit?tab=t.0#heading=h.6qdq66n7hr47)
- **Spanish:** [LE Resumen ejecutivo](https://docs.google.com/document/d/1bCfCDMTlREiOwrVbwFCV9vXM56OKT5OTxfHDjMfdFdo/edit?tab=t.0)
- **EWG 3.1 page:** [Livelihoods & Economies](ewg31_le.html)
- **Goalkeeper page:** [Livelihoods and economies](le.html)

:::

::: {.callout-note collapse="true"}
# Next steps

- *To be added from EWG feedback.*
:::

# ![](figs/sense-of-place.svg){width=8% height=8%} Sense of Place {#sp}

```{r}
#| echo: false
#| warning: false
#| message: false
sp_cs <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/final_cs_scores/SP_Scores_1.0_df_091126.csv"),
  show_col_types = FALSE
) %>%
  prep_cs()
goal_plotly(sp_cs, "Sense of Place current status")
```

::: {.callout-tip collapse="true"}
# About this goal

The Sense of Place (SP) goal — *Sentido de Pertenencia* — measures how well we are sustainably caring for the marine and coastal places and species that provide heritage, shared memory, and identity for Gulf communities. Specific data sources and reference points are described in detail in the executive summary.

- **English:** [SP Executive Summary](https://docs.google.com/document/d/1qxuYHPqFbGprBSflfiq00w3IlXs-9VjmsxYXM4O3sgY/edit?tab=t.0)
- **Spanish:** [SP Resumen ejecutivo](https://docs.google.com/document/d/1Bs14irzJuaMGkqVljkC5fpJot0FZvwz0l0jlmNeV_fc/edit?tab=t.0)
- **EWG 3.2 page:** [Sense of Place](ewg32_sp.html)
- **Goalkeeper page:** [Sense of Place](sp.html)

:::

::: {.callout-note collapse="true"}
# Next steps

- *To be added from EWG feedback.*
:::

# ![](figs/artisanal-fishing-opportunities.svg){width=8% height=8%} Access Opportunities {#ao}

```{r}
#| echo: false
#| warning: false
#| message: false
ao_cs <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/final_cs_scores/AO_Scores_1.0_df_Sep112026.csv"),
  show_col_types = FALSE
) %>%
  prep_cs()
goal_plotly(ao_cs, "Access Opportunities current status")
```

*Sophia will move Erica's Access Opportunities current-status plot into `figs/` before Friday.*

::: {.callout-tip collapse="true"}
# About this goal

Access Opportunities (AO), formerly Artisanal Opportunities, measures whether members of coastal communities — including small-scale fishing communities and *pueblos originarios* — have access to marine resources that allow them to thrive. Specific data sources and reference points are described in detail in the executive summary.

- **English:** [AO Executive Summary](https://docs.google.com/document/d/1RyoGspGrQRQ4deif7Ybm3k3e39EG2X9dBnpket8ITtk/edit?tab=t.0)
- **Spanish:** [AO Resumen Ejecutivo](https://docs.google.com/document/d/146yfYH7c9UaSo3PGg5DI2u7YRK1iDVxIQeUxPckOO0Y/edit?tab=t.0)
- **EWG 3.2 page:** [Access Opportunities](ewg32_ao.html)
- **Goalkeeper page:** [Access Opportunities](ao.html)

:::

::: {.callout-note collapse="true"}
# Next steps

- *To be added from EWG feedback.*
:::

# ![](figs/coastal-protection.svg){width=8% height=8%} Habitat Services {#hs}

```{r}
#| echo: false
#| warning: false
#| message: false
hs_cs <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/final_cs_scores/HS_Scores_1.0_df_Sep112026.csv"),
  show_col_types = FALSE
) %>%
  prep_cs()
goal_plotly(hs_cs, "Habitat Services current status")
```

*The Habitat Services current-status plot is not ready yet. Erica can make it before Monday, and Sophia will move it into `figs/` once it is available.*

::: {.callout-tip collapse="true"}
# About this goal

Habitat Services (HS) combines Coastal Protection and Carbon Storage. It measures the ability of coastal habitats to protect people and property from flooding and erosion, and to sequester and store atmospheric carbon. Specific data sources and reference points are described in detail in the executive summary.

- **English:** [HS Executive Summary](https://docs.google.com/document/d/1mro2xQ3Q1TRTs5apmPCkJcBdkZCdjq8U3asCaNsrY9k/edit?tab=t.0)
- **Spanish:** [HS Resumen Ejecutivo](https://docs.google.com/document/d/1xJIrO15HS-U_P684PrhH_FTo9zjJgJBQiAhvTrkYHXs/edit?tab=t.0)
- **EWG 3.2 page:** [Habitat Services](ewg32_hs.html)
- **Goalkeeper page:** [Habitat services (CP & CS)](cp_cs.html)

:::

::: {.callout-note collapse="true"}
# Next steps

- *To be added from EWG feedback.*
:::

# ![](figs/biodiversity.svg){width=8% height=8%} Biodiversity {#bd}

```{r}
#| echo: false
#| warning: false
#| message: false
bd_cs <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/final_cs_scores/BD_Scores_1.0_df_Sep112026.csv"),
  show_col_types = FALSE
) %>%
  prep_cs()
goal_plotly(bd_cs, "Biodiversity current status")
```

*Sophia will move Erica's Biodiversity current-status plot into `figs/` before Friday.*

::: {.callout-tip collapse="true"}
# About this goal

The Biodiversity (BD) goal measures the current status of key biomes as proxies for species diversity, abundance, and ecosystem function. Specific data sources and reference points are described in detail in the executive summary.

- **English:** [BD Executive Summary](https://docs.google.com/document/d/19N-MjE9iucG71N6NUDzUimqSB1tA-4XCVWj9SNXyy6w/edit?tab=t.0)
- **Spanish:** [BD Resumen Ejecutivo](https://docs.google.com/document/d/1zERminsNjyKGoNBz7TqWHqsOSnNl3bKdDzHeIr4VIaQ/edit?tab=t.0)
- **EWG 3.3 page:** [Biodiversity](ewg33_bd.html)
- **Goalkeeper page:** [Biodiversity](bd.html)

:::

::: {.callout-note collapse="true"}
# Next steps

- *To be added from EWG feedback.*
:::

# ![](figs/clean-waters.svg){width=8% height=8%} Clean Waters {#cw}

The overall CW score is the average of Eutrophication, Pathogens, Chemicals, and Marine Debris. A score of 100 is the ideal clean-waters state. Open the component scores below to see what is pulling a region up or down. Specific data sources and reference points are described in detail in the executive summary.

```{r}
#| echo: false
#| warning: false
#| message: false

cw_root <- "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025"

cw_pick <- function(...) {
  paths <- unlist(list(...), use.names = FALSE)
  hit <- paths[file.exists(paths)]
  if (length(hit) == 0) {
    stop("None of these CW files exist: ", paste(paths, collapse = "; "))
  }
  hit[[1]]
}

cw_normalize_rgn <- function(x) {
  x <- as.character(x)
  ifelse(grepl("^Region_", x), x, paste0("Region_", x))
}

cw_prep <- function(df, score_col) {
  df %>%
    mutate(
      rgn_id = cw_normalize_rgn(rgn_id),
      year = as.numeric(year),
      current_status = as.numeric(.data[[score_col]])
    ) %>%
    add_rgn_labels()
}

cw_eut <- read_csv(
  cw_pick(file.path(cw_root, "eutrophication/output/eutrophication_cs.csv")),
  show_col_types = FALSE
) %>%
  dplyr::select(rgn_id, current_status) %>%
  group_by(rgn_id) %>%
  mutate(year = 2017) %>%
  complete(year = 2017:2025) %>%
  fill(current_status, .direction = "down") %>%
  ungroup() %>%
  cw_prep("current_status")

cw_pat <- read_csv(
  file.path(cw_root, "pathogens/output", "pathogen_cs_2018_2025.csv"),
  show_col_types = FALSE
) %>%
  dplyr::select(rgn_id, year, current_status) %>%
  mutate(
    year = as.numeric(year),
    current_status = suppressWarnings(as.numeric(current_status))
  ) %>%
  filter(year != 2020) %>%
  cw_prep("current_status")

cw_chm <- read_csv(
  cw_pick(
    file.path(cw_root, "chemicals/output/chemical_cs.csv"),
    file.path(cw_root, "chemicals/chemical_cs.csv")
  ),
  show_col_types = FALSE
) %>%
  dplyr::select(rgn_id, year, chemical_cs) %>%
  group_by(rgn_id) %>%
  complete(year = 2017:2025) %>%
  fill(chemical_cs, .direction = "down") %>%
  ungroup() %>%
  cw_prep("chemical_cs")

cw_mar_raw <- read_csv(
  cw_pick(
    file.path(cw_root, "marine_debris/ghost_gear_risk.csv"),
    "/home/shares/ohi/OHI_GOC/goal_prep/_pressures/vessels/output/ghost_gear_risk.csv"
  ),
  show_col_types = FALSE
)

if ("pressure" %in% names(cw_mar_raw)) {
  cw_mar <- cw_mar_raw %>%
    mutate(current_status = (1 - pressure) * 100) %>%
    dplyr::select(rgn_id, year, current_status)
} else {
  score_col <- intersect(c("current_status", "PLA", "final_score"), names(cw_mar_raw))[1]
  cw_mar <- cw_mar_raw %>%
    dplyr::select(rgn_id, year, current_status = all_of(score_col))
}

cw_mar <- cw_prep(cw_mar, "current_status")

cw_overall <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/final_cs_scores/CW_Scores_1.0_df_091126.csv"),
  show_col_types = FALSE
) %>%
  prep_cs()

goal_plotly(cw_overall, "Clean Waters current status")
```

- **English:** [CW Executive Summary](https://docs.google.com/document/d/1VRxaDW9YbCoammwVcZut07vNqpNpUcT8fo_kM2V47is/edit?tab=t.0)
- **Spanish:** [CW Resumen Ejecutivo](https://docs.google.com/document/d/1wq48kxZkAXhOMZS6kEPYYA-u9jZ2erKhexnGE0nDSTM/edit?tab=t.0)
- **EWG 3.3 page:** [Clean Waters](ewg33_cw.html)
- **Goalkeeper page:** [Clean waters](cw.html)

::: {.callout-note collapse="true"}
# Component scores

Specific data sources and reference points for each component are described in detail in the executive summary.

This is the Eutrophication component, which evaluates excess nitrogen reaching the coast from wastewater, agriculture, livestock, and shrimp aquaculture, with a score of 100 meaning zero nitrogen pollution.

```{r}
#| echo: false
#| warning: false
#| message: false
goal_plotly(cw_eut, "CW Eutrophication")
```

This is the Pathogens component, which evaluates *Enterococcus* at Playas Limpias beaches, with a score of 100 meaning waters meet Mexico’s beach suitability standard.

```{r}
#| echo: false
#| warning: false
#| message: false
goal_plotly(cw_pat, "CW Pathogens")
```

This is the Chemicals component, which evaluates pesticide, pharmaceutical, hazardous-waste, untreated-wastewater, and vessel-hydrocarbon risk, with a score of 100 meaning no chemical pressure relative to the highest levels observed in the Gulf.

```{r}
#| echo: false
#| warning: false
#| message: false
goal_plotly(cw_chm, "CW Chemicals")
```

This is the Marine Debris component, which evaluates the risk of abandoned fishing gear (ghost gear), with a score of 100 meaning no ghost-gear pressure.

```{r}
#| echo: false
#| warning: false
#| message: false
goal_plotly(cw_mar, "CW Marine Debris")
```
:::

::: {.callout-note collapse="true"}
# Next steps and questions for Alejandro

**Next steps (from EWG 3.3)**

- *To be added from EWG feedback.*

**Questions for Alejandro**

- Do these scores seem right to you?
- Do you have any major questions or concerns you would like to discuss?
- How do you feel about the eutrophication methods? Does the low score in the Western Upper Gulf (short coastline, large livestock area) seem believable, and should we rerun this for multiple years?
- For indicators that use the highest observed Gulf pressures as a score of zero, does that feel representative of the Gulf?
:::

# ![](figs/natural-products.svg){width=8% height=8%} Natural Products {#np}

```{r}
#| echo: false
#| warning: false
#| message: false
np_cs <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/final_cs_scores/NP_Scores_1.0_df_Sep112026.csv"),
  show_col_types = FALSE
) %>%
  prep_cs()
goal_plotly(np_cs, "Natural Products current status")
```

*Sophia will move Erica Natural Products current-status plot into `figs/` before Friday.*

::: {.callout-tip collapse="true"}
# About this goal

The Natural Products (NP) goal measures the amount, benefits, and sustainability of legally harvested marine resources produced in the Gulf of California for cultural or commercial purposes that are not used for food. Specific data sources and reference points are described in detail in the executive summary.

- **English:** [NP Executive Summary](https://docs.google.com/document/d/1QaEn1ucoh_ToqsrguQGfGqRSK3VjT9GBk6vqlfVxBtg/edit?tab=t.0)
- **Spanish:** [NP Resumen Ejecutivo](https://docs.google.com/document/d/1ymx6DLhzzLjXM-Ha9JMlPA39NenC9NNrYVTKLZ_Vfpc/edit?tab=t.0)
- **EWG 3.3 page:** [Natural Products](ewg33_np.html)
- **Goalkeeper page:** [Natural products](np.html)

:::

::: {.callout-note collapse="true"}
# Next steps

- *To be added from EWG feedback.*
:::
