---
title: "Pharmaceutical sources (mg) choropleths"
format:
  html:
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

Choropleths of **mg load by OHI region** for each pharmaceuticals source, **before** aggregation into current status:

1. Human wastewater (WWTP)
2. Livestock antimicrobials
3. Shrimp farming antimicrobials

```{r}
library(tidyverse)
library(sf)
library(here)
library(scales)
library(grid)

pick_existing <- function(...) {
  paths <- unlist(list(...), use.names = FALSE)
  hit <- paths[file.exists(paths)]
  if (length(hit) == 0) NA_character_ else hit[[1]]
}

normalize_rgn_id <- function(x) {
  x %>% as.character() %>% str_squish() %>% str_replace_all(" ", "_")
}

rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

cw_root <- "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025"
local_pharm <- here::here("cw_mazu/v2025/chemicals/pharmaceuticals")

# Paths follow script4 on Mazu: .../v2025/pharmaceuticals/...
# Local copy often lives under chemicals/pharmaceuticals/
wwtp_path <- pick_existing(
  file.path(cw_root, "pharmaceuticals/humans_output/wwtp_mg_rgnid_pharma.csv"),
  file.path(cw_root, "chemicals/pharmaceuticals/humans_output/wwtp_mg_rgnid_pharma.csv"),
  file.path(local_pharm, "humans_output/wwtp_mg_rgnid_pharma.csv")
)

livestock_path <- pick_existing(
  file.path(cw_root, "pharmaceuticals/livestock_output/watershed_antimicrobials_livestock.csv"),
  file.path(cw_root, "chemicals/pharmaceuticals/livestock_output/watershed_antimicrobials_livestock.csv"),
  file.path(local_pharm, "livestock_output/watershed_antimicrobials_livestock.csv")
)

shrimp_path <- pick_existing(
  file.path(cw_root, "pharmaceuticals/shrimp_output/shrimp_mg_pharmaceuticals_rgnid.csv"),
  file.path(cw_root, "chemicals/pharmaceuticals/shrimp_output/shrimp_mg_pharmaceuticals_rgnid.csv"),
  file.path(local_pharm, "shrimp_output/shrimp_mg_pharmaceuticals_rgnid.csv")
)

rgn_path <- pick_existing(
  "/home/shares/ohi/OHI_GOC/spatial/OHI_regions/goc_ohi_rgns.shp",
  "/home/shares/ohi/OHI_GOC/spatial/ohi_regions/goc_ohi_rgns.shp",
  here::here("spatial/ohi_regions/goc_ohi_rgns.shp")
)

stopifnot(!is.na(wwtp_path), !is.na(livestock_path), !is.na(shrimp_path), !is.na(rgn_path))

regions <- st_read(rgn_path, quiet = TRUE) %>%
  mutate(rgn_id = normalize_rgn_id(rgn_id))
if ("location" %in% names(regions)) {
  regions <- regions %>% filter(location == "marine" | is.na(location))
}

read_source_mg <- function(path, mg_col, source_name) {
  read_csv(path, show_col_types = FALSE) %>%
    mutate(
      rgn_id = normalize_rgn_id(rgn_id),
      year = as.numeric(year),
      mg = as.numeric(.data[[mg_col]]),
      source = source_name
    ) %>%
    select(rgn_id, year, mg, source)
}

wwtp <- read_source_mg(wwtp_path, "total_pharma_mg", "Human wastewater (WWTP)")
livestock <- read_source_mg(livestock_path, "mg_antimicrobials", "Livestock antimicrobials")
shrimp <- read_source_mg(shrimp_path, "mg_antimicrobials", "Shrimp farming antimicrobials")

# Reverse of CW score palette: blue = least mg, red = most mg
mg_palette <- rev(c(
  "#8a0032", "#A50026", "#A50026", "#D73027", "#F46D43",
  "#FDAE61", "#FEE090", "#FEE090", "#ABD9E9", "#4575B4", "#313695"
))

latest_map <- function(df, title) {
  y <- max(df$year[!is.na(df$mg)], na.rm = TRUE)
  latest <- df %>%
    filter(year == y, !is.na(mg)) %>%
    select(rgn_id, year, mg)
  map_dat <- regions %>%
    left_join(latest, by = "rgn_id") %>%
    mutate(rgn_label = unname(rgn_labels[rgn_id]))

  mg_pos <- latest$mg[latest$mg > 0 & is.finite(latest$mg)]
  if (length(mg_pos) == 0) {
    fill_limits <- c(1, 10)
    fill_breaks <- c(1, 10)
  } else {
    fill_limits <- range(mg_pos, na.rm = TRUE)
    # evenly spaced breaks in log10 space; keep few labels so they do not collide
    log_rng <- log10(fill_limits)
    fill_breaks <- unique(10^pretty(log_rng, n = 4))
    fill_breaks <- fill_breaks[
      fill_breaks >= fill_limits[1] * 0.999 & fill_breaks <= fill_limits[2] * 1.001
    ]
    if (length(fill_breaks) < 2) fill_breaks <- fill_limits
  }

  p <- ggplot() +
    geom_sf(data = map_dat, aes(fill = mg), color = "black", linewidth = 0.25) +
    scale_fill_gradientn(
      colours = mg_palette,
      trans = "log10",
      limits = fill_limits,
      breaks = fill_breaks,
      labels = label_scientific(digits = 1),
      na.value = "grey90",
      name = "mg\n(red = highest)",
      guide = guide_colorbar(
        title.position = "top",
        barwidth = unit(8, "cm"),
        barheight = unit(0.45, "cm"),
        nbin = 100
      )
    ) +
    labs(
      title = title,
      subtitle = paste0("mg load by region | single year: ", y, " | red = most, blue = least")
    ) +
    theme_void(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      legend.position = "bottom",
      legend.title = element_text(hjust = 0.5),
      legend.text = element_text(size = 8)
    )

  list(year = y, data = latest, plot = p)
}

cat("CW root:", cw_root, "\n")
cat("WWTP:", wwtp_path, "\n")
cat("Livestock:", livestock_path, "\n")
cat("Shrimp:", shrimp_path, "\n")
cat("Regions:", rgn_path, "\n")
```

## Human wastewater (mg)

```{r}
#| fig-width: 7
#| fig-height: 7.5
wwtp_map <- latest_map(wwtp, "Human wastewater pharmaceuticals (mg)")
wwtp_map$plot
```

```{r}
wwtp_map$data %>%
  mutate(rgn_label = unname(rgn_labels[rgn_id])) %>%
  arrange(desc(mg)) %>%
  mutate(mg = comma(mg, accuracy = 1)) %>%
  select(rgn_id, rgn_label, year, mg) %>%
  knitr::kable()
```

## Livestock antimicrobials (mg)

```{r}
#| fig-width: 7
#| fig-height: 7.5
livestock_map <- latest_map(livestock, "Livestock antimicrobials (mg)")
livestock_map$plot
```

```{r}
livestock_map$data %>%
  mutate(rgn_label = unname(rgn_labels[rgn_id])) %>%
  arrange(desc(mg)) %>%
  mutate(mg = comma(mg, accuracy = 1)) %>%
  select(rgn_id, rgn_label, year, mg) %>%
  knitr::kable()
```

## Shrimp farming antimicrobials (mg)

```{r}
#| fig-width: 7
#| fig-height: 7.5
shrimp_map <- latest_map(shrimp, "Shrimp farming antimicrobials (mg)")
shrimp_map$plot
```

```{r}
shrimp_map$data %>%
  mutate(rgn_label = unname(rgn_labels[rgn_id])) %>%
  arrange(desc(mg)) %>%
  mutate(mg = comma(mg, accuracy = 1)) %>%
  select(rgn_id, rgn_label, year, mg) %>%
  knitr::kable()
```

## Side-by-side latest mg (all sources)

Uses each source's own latest year (they do not all end in the same year).

```{r}
bind_rows(
  wwtp_map$data %>% mutate(source = "Human wastewater"),
  livestock_map$data %>% mutate(source = "Livestock"),
  shrimp_map$data %>% mutate(source = "Shrimp")
) %>%
  mutate(rgn_label = unname(rgn_labels[rgn_id])) %>%
  select(source, rgn_id, rgn_label, year, mg) %>%
  pivot_wider(names_from = source, values_from = c(year, mg)) %>%
  knitr::kable()
```
