---
title: " "
format: html
toc: true
---


## Fishery Production Data in the Gulf of California

Mexico's commercial fishing industry is regulated by CONAPESCA (Comision Nacional de Acuacultura y Pesca), a specialized agency operating under SADER (Secretaría de Agricultura y Desarrollo Rural), Mexico's agriculture ministry. CONAPESCA oversees the development of sustainable fishing practices and implements policies to enhance the competitiveness of Mexico's fishing sector.

### Commercial Fishing Requirements

All commercial fishing operations in Mexico must obtain licenses that require reporting:

-  Landing location
-  Species caught
-  Transportation method
-  Catch volume
-  Presence of any protected species (en veda)

### Interpretation of Fishery Production

Live weight (also called "round weight" or "whole weight") refers to the total mass of fish when they're first caught, before any processing (e.g. gutting, heading, filleting). It represents the fish in their complete, original form right out of the water. Measuring in live weight helps standardize catch reporting across different fishing operations, track the total biomass removed from marine ecosystems, monitor fishing quotas and stock assessments, and compare catches across different species and regions.

Below, you can select one species for both the live weight and revenue over time within the Gulf of California. When comparing the two, consider:

-  Different species are associated with different prices per kilogram
-  Market conditions and supply chains influence value over time
-  Quality and freshness impact price, which could be impaired in years where the stock population's health is at risk
-  Processing costs affect final revenue
-  These are only marine, wild-caught stocks

Together, these metrics help explore both the ecological and economic sustainability of fishing operations.


```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(magrittr)
library(here)
library(tictoc)
library(kableExtra)
library(plotly)
library(viridis)
library(readr)
library(purrr)

# ---- sources! ----
# source(here("workflow", "R", "common.R")) # file creates objects to process data
## set the mazu and neptune data_edit share based on operating system
dir_M             <- c('Windows' = '//mazu.nceas.ucsb.edu/ohi',
                       'Darwin'  = '/Volumes/ohi',    ### connect (cmd-K) to smb://mazu/ohi
                       'Linux'   = '/home/shares/ohi')[[ Sys.info()[['sysname']] ]]

# ---- set year and file path info ----
current_year <- 2025 # Update this in the future!!
version_year <- paste0("v",current_year)
data_dir_version_year <- paste0("d", current_year)

# ---- data directories ----

# raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "OHI_GOC", "_raw_data")

# fishery production per office raw data directory
fish_office_dir <- here(raw_data_dir, "CONAPESCA", data_dir_version_year, "fishery_offices")

# annual statistics raw data directory
stat_dir <- here(raw_data_dir, "CONAPESCA", data_dir_version_year, "statistical_yearbook_fishery_aquaculture")

# final output dir
output_dir <- here("FIS", version_year, "output")
```


```{r csv, echo=FALSE, warning=FALSE, message=FALSE}
# read in habitat_marine_species_summary.csv
habitat_marine_species_summary <- read_csv(here::here(fish_office_dir, "int","habitat_marine_species_summary.csv"))
```
### Total Catch in Gulf of California (2006-2022)

```{r total_catch, echo=FALSE, warning=FALSE, message=FALSE}
# summarize the kg and revenue across all fishery offices in the GoC for each year for ONLY marine species
total_fish_wild_goc_summary <- habitat_marine_species_summary %>% 
  group_by(ejercicio_fiscal) %>% 
  summarize(
    agg_live_weight = sum(agg_live_weight),
    agg_revenue = sum(agg_revenue)
  )

# plot live weight interactively using plot_ly
total_live_weight_plotly <- total_fish_wild_goc_summary %>%
  plot_ly(x = ~ejercicio_fiscal, y = ~agg_live_weight, 
          type = "scatter", mode = "lines", line = list(color = "deepskyblue4")) %>%
  layout(title = " ",
         xaxis = list(title = "Year"),
         yaxis = list(title = "Live Weight (kg)"))
total_live_weight_plotly
```

### Total Revenue in in Gulf of California (2006-2022)

```{r total_rev, echo=FALSE, warning=FALSE, message=FALSE}
# save plot
# htmlwidgets::saveWidget(total_live_weight_plotly, file = "output/total_live_weight_2006_2022_plot.html")

# plot revenue interactively using plot_ly
total_revenue_plotly <- total_fish_wild_goc_summary %>%
  plot_ly(x = ~ejercicio_fiscal, y = ~agg_revenue, 
          type = "scatter", mode = "lines", line = list(color = "darkgreen")) %>%
  layout(title = " ",
         xaxis = list(title = "Year"),
         yaxis = list(title = "Revenue (pesos)"))
total_revenue_plotly

# save plot
# htmlwidgets::saveWidget(total_revenue_plotly, file = "output/total_revenue_2006_2022_plot.html")
```

### Annual Live Weight of Marine Species Caught in Gulf of California (2006-2022)

```{r live_weight, echo=FALSE, warning=FALSE, message=FALSE}
# plot live weight interactively using plot_ly
marine_live_weight_plotly <- habitat_marine_species_summary %>%
  plot_ly(
    x =  ~ ejercicio_fiscal,
    y =  ~ agg_live_weight,
    type = "scatter",
    mode = "lines",
    # set hover text for better readability
    text =  ~ paste(
      "Species:",
      nombre_principal,
      "<br>Year:",
      ejercicio_fiscal,
      "<br>Live Weight:",
      agg_live_weight
    ),
    hoverinfo = "text",
    transforms = list(
      list(
        type = 'filter',
        target =  ~ nombre_principal,
        operation = '=',
        value = unique(habitat_marine_species_summary$nombre_principal)[1] 
      )
    ),
    line = list(color = "deepskyblue4") 
  ) %>%
  layout(
    title = " ",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Live Weight (kg)"),
    updatemenus = list(
      list(
        type = 'dropdown',
        active = 0, # default is first group
        buttons = lapply(
          unique(habitat_marine_species_summary$nombre_principal), 
          function(species) {
            list(
              method = "restyle",
              args = list(
                "transforms[0].value", 
                species
              ),
              label = species 
            )
          }
        )
      )
    )
  )

marine_live_weight_plotly
```

### Annual Revenue from Marine Species in Gulf of California (2006-2022)

```{r revenue, echo=FALSE, warning=FALSE, message=FALSE}
# plot revenue interactively using plot_ly
marine_revenue_plotly <- habitat_marine_species_summary %>%
  plot_ly(
    x =  ~ ejercicio_fiscal,
    y =  ~ agg_revenue,
    type = "scatter",
    mode = "lines",
    # set hover text for better readability
    text =  ~ paste(
      "Species:",
      nombre_principal,
      "<br>Year:",
      ejercicio_fiscal,
      "<br>Revenue:",
      agg_revenue
    ),
    hoverinfo = "text",
    transforms = list(
      list(
        type = 'filter',
        target =  ~ nombre_principal,
        operation = '=',
        value = unique(habitat_marine_species_summary$nombre_principal)[1] 
      )
    ),
    line = list(color = "darkgreen") # changing color
  ) %>%
  layout(
    title = " ",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Revenue (pesos)"),
    updatemenus = list(
      list(
        type = 'dropdown',
        active = 0, 
        buttons = lapply(
          unique(habitat_marine_species_summary$nombre_principal), 
          function(species) {
            list(
              method = "restyle",
              args = list(
                "transforms[0].value", 
                species 
              ),
              label = species 
            )
          }
        )
      )
    )
  )

marine_revenue_plotly
```


## Data Source

**Reference**: https://datos.gob.mx/busca/dataset/produccion-pesquera 

**Downloaded**: February 11, 2025

**Description**: Fisheries Production

**Time range**: 2006 - 2022

**Format**: .csv

**File location**: `Mazu:/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/fishery_offices`

