---
title: " "
format: html
toc: true
---

## Temporary accommodation 

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
# refer to ohi_goc_prep/LSP/v2025/lsp_wdpa_data_prep.qmd for full data prep
library(sf)
library(tidyverse)
library(mapview)
library(leafpop)
library(here)
```

The ability to visit culturally or aesthetically pleasing places with healthy coastal ecosystems, diverse wildlife, and opportunities for exploration and recreation contributes to the well-being of many people and is an important driver of coastal economies. 

An obvious way to assess tourism in an area would be to directly count the number of tourists visiting each location. Tourism data are available, but the data we found did not describe where the tourists were traveling to within Mexico and only included foreign visitors.

Another way to measure tourism is through the presence of businesses that provide accommodations and food. These data are from [INEGI](https://www.inegi.org.mx/app/descarga/?ti=6). Location data is provided for temporary accommodations and food and beverage services (although we only present accommodations here). There is also a category describing the number of employees. 

We limited the accommodations to those within 50km of the coast (this could easily be altered).

Categories of accommodations include:

- Boarding houses and guesthouses (n = 118) 
- Cabins, villas, and similar accommodations (n = 352) 
- Camps and recreational hostels (n = 67) 
- Furnished apartments and houses with hotel services (n = 159) 
- Hotels with integrated services (n = 686) 
- Hotels without integrated services (n = 717) 
- Motels (n = 234)

### Interactive Map

The map below shows the locations of available accommodations. The size reflects the number of employees (which is a category). Points are colored by the number of employees.

```{r hotels, echo=FALSE, warning=FALSE, message=FALSE}

# read in data
hotels <- 
  st_read("https://raw.githubusercontent.com/OHI-GOC/ohi_goc_prep/refs/heads/gh-pages/TR/data/accommodations/accommodations_50km_2024.geojson", quiet=TRUE)
# make mapview interactive map
mapview(hotels,
                             zcol = "rank_ocu",  # color by IUCN category
                             layer.name = "Accommodations",
                             map.types = c("OpenStreetMap", "Esri.WorldShadedRelief", "Esri.WorldImagery"),
                             alpha.regions = 0.8,
                             popup = popupTable(hotels, zcol = c("nom_estab", "est_type", "per_ocu"))) 
```

## Download data

Processed data are available [here](https://github.com/OHI-GOC/ohi_goc_prep/tree/gh-pages/TR/data/accommodations). 

------------------------------------------------------------------------
**Reference**: Instituto Nacional de Estadística y Geografía (INEGI), Datos geoespaciales, https://www.inegi.org.mx/app/descarga/?ti=6 (accessed 23 April 2025).

**Downloaded**: March 5, 2025

**Description**: Latitude and Longitudes of temporary accommodations and food establishments.

**Time range**: 11/2024 (although I believe previous )

**Format**: CSV files
